---
title: "dplyr & pipes"
author: "Ellen Dymit"
date: "11/4/2020"
output: pdf_document
---

## dplyr

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

familiarizing yourself with dplyr will change your life! 

the dplyr package exists to make tabular data manipulation cleaner and easier

the data we're working with today is in the tidyverse package. 

```{r}
starwars

glimpse(starwars)

#starwars <- dplyr::tbl_df(starwars)
```


dplyr is very simple. the package is built around 5 verbs. 

SELECT
FILTER
ARRANGE
MUTATE
SUMMARISE

These verbs make up the majority of the data manipulation you tend to do. 
The first element in all 5 functions is your dataframe.* (Note this, it'll be important later)

### select
Select certain columns of data.

```{r}

select(starwars, name, species, homeworld) # note that you can reorder the columns here

select(starwars, name, hair_color:birth_year) # specify a "run" of columns.

select(starwars, -vehicles, -starships) # exclude columns

select(starwars, name, contains("color"))
select(starwars, name, ends_with("color")) # starts_with

select(starwars, 1:5)

```

### filter
Subset by rows

```{r}

filter(starwars, height > 100)
filter(starwars, mass < 40)[ , c(1:3, 11)] # you can use any valid logical statement here

filter(starwars, !is.na(hair_color))

filter(starwars, species == "Droid")[ , 1:3]
filter(starwars, species == "Human" & sex == "female")[ , c(1, 4:6)]

```

### arrange
Order your rows

```{r}

arrange(starwars, name)[ , c(1, 4:6)] # default is alphabetical

arrange(starwars, desc(height))[ , c(1, 2, 10:11)]

arrange(starwars, species, name)[ , c(1, 2, 10:11)]

```

### mutate
Add new columns

```{r}

Loblolly
mutate(Loblolly, height_feet = height / 3.281)
mutate(Loblolly, height_rank = height / mean(height), height_rank = round(height_rank, 3))

iris
mutate(iris, Sepal.Size = Sepal.Length * Sepal.Width)
mutate(iris,
       Sepal.Size = Sepal.Length * Sepal.Width,
       Petal.Size = Petal.Length * Petal.Width
       )[ , 5:7]

```

### summarize
calculate summary statistics

```{r}

Loblolly
summarize(Loblolly, mean_height = mean(height))

# summarize is more useful when combined with group_by

head(summarize(group_by(Loblolly, age),
  mean_height = mean(height)))

```


## pipes

this is the pipe operator in R: %>%

(most important reason to use pipes is that its just fun to type fast while holding down shift key)

it lives in the 'magrittr' package, which comes in the tidyverse

pipes take the output of the first function and feet it into the subsequent one
think of it as the word "then"

pipes are how you design your desired data frame without creating so many intermediate objects 

the pipe operator truly shines when you start chaining several dplyr functions with it

```{r, error=TRUE}

x <- rnorm(10)

x %>% mean 
mean(x)

starwars

starwars %>%
  filter(sex == "female") %>%
  select(name, species, homeworld)

# can anyone tell me why this returns an error?
starwars %>%
  select(name, species, homeworld) %>%
  filter(sex == "female")

# Q: which species is the strongest / weakest
starwars %>%
  filter(species != "Droid") %>%
  mutate(power = mass * height) %>%
  group_by(species) %>%
  summarize(avg_strength = mean(power)) %>%
  filter(!is.na(avg_strength)) %>%
  arrange(desc(avg_strength)) %>%
  mutate(rank = row_number()) %>%
  select(rank, species)
  
```


## piping into ggplot


```{r}

iris.plot <- iris %>% # variable name and assignment on the same line is best style, imo
  mutate(Sepal.Size = Sepal.Length * Sepal.Width,
       Petal.Size = Petal.Length * Petal.Width) %>%
  ggplot(., aes(Petal.Size, Sepal.Size, color = Species)) +
  geom_point()
iris.plot

```

## reshaping data

```{r}

# gather
starwars1 <- starwars %>%
  gather(key = "feature", value = "color", hair_color, skin_color, eye_color # or 4:6
         ) %>%
  select(name, feature, color) %>%
  filter(!is.na(color))
starwars1

# spread
starwars1 %>%
  spread(key = "feature", value = "color")

# unite
starwars %>%
  filter(species != "Droid") %>%
  unite("identity", species, sex, sep = " ") %>% # default separator is _
  select(name, identity)

# separate
starwars %>%
  filter(species != "Droid") %>%
  separate(name, c("first_name", "last_name"), sep = " ")

```
